% !TEX root = main.tex

\thispagestyle{plain}
\begin{center}
    \Large
    \textbf{Efficient Synchronization of Recursively Partitionable Data Structures}
       
    \vspace{0.9cm}
    \textbf{Abstract}
\end{center}

Given two nodes in a distributed system, each of them holding a data structure, one or both of them might need to update their local replica based on the data available at the other node.
An efficient solution should avoid redundantly sending data to a node which already holds it.

We give conceptually simple yet asymptotically efficient probabilistic solutions based on recursively exchanging fingerprints for data structures of exponentially decreasing size, obtained by recursively partitioning the data structures.
We apply the technique to sets, maps and radix trees.
For data structures containing $n$ items, this leads to $\complexity{log(n)}$ round-trips.
We give a scheme by which the fingerprints can be computed in $\complexity{log(n)}$ time, based on an auxiliary data structure which requires $\complexity{n}$ space and which can be updated to reflect changes to the underlying data structure in $\complexity{log(n)}$ time.

To minimize the number of round-trips, the technique requires up to $\complexity{n}$ space per synchronization session.
It can be adapted to require only a bounded amount of memory, which is essential for robust, scalable implementations.
While this increases the worst-case number of round-trips, it guarantees continuous progress, even in adversarial environments.